---
title: "Publications"
description: |
  Recent publications and conference presentations. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r write_refs, include = FALSE, cache = FALSE}
library(RefManageR)
# Grab bibTeX using DOIs, and save to pubs.bib
RefManageR::GetBibEntryWithDOI(
  unique(
    c(readLines("pub_doi.txt"))
  ), 
  temp.file = "pubs.bib", 
  delete.file = FALSE)
# Manual bibTeX for articles without DOIs
ref <- RefManageR::ReadBib("pub_nodoi.bib")
WriteBib(bib = ref, 
         file = "pubs.bib", 
         append = TRUE)
```

```{r, load_refs, include = FALSE, cache = FALSE}
library(RefManageR)
# Read "pubs.bib"
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE, 
           no.print.fields = "url")
myBib <- ReadBib(here::here("pubs.bib"), check = FALSE)
```

## Peer-Reviewed Articles (Selected)

### 2022

```{r, results = "asis"}
myBib[date = "2022"]
```

### 2021

```{r, results = "asis"}
myBib[date = "2021"]
```

### 2020

```{r, results = "asis"}
myBib[date = "2020"]
```


### 2019

```{r, results = "asis"}
myBib[date = "2019"]
```

## Conference Presentations

### 2021

Zhang, G., Yue, H., & Lai, M. H. C. (2021, July 19-23). *A systematic review of measurement invariance research of the CES-D across gender* [Poster presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Zhang, Y. & Lai, M. H. C. (2021, July 19-23). *Classification Accuracy of multidimensional tests: Quantifying the impact of noninvariance* [Oral presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Tse, W. W. & Lai, M. H. C. (2021, July 19-23). *Incorporating uncertainty in design parameters: A hybrid classical-Bayesian power analysis approach for two-level cluster randomized trials* [Oral presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Lai, M. H. C., Ji, F., & Chen, S. (2021, July 19-23). *Obtaining reliability for daily diary data using multilevel factor analysis* [Oral presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Hsiao, Y., Kwok, O., & Lai, M. H. C. (2021, April 8-12). *Improving estimation accuracy of the tnteraction effects in models with composite scores* [Structured poster session]. AERA Annual Meeting, Orlando, FL, United States. https://convention2.allacademic.com/one/aera/aera21

Lai, M. H. C. & Hsiao, Y. (2021, April). *Two-stage path analysis with definition variables: A general framework for modeling with categorical indicators* [Structured Poster Session]. AERA Annual Meeting, Orlando, FL, United States. 

Wen, L., & Lai, M. H. C. (2021, April 8-12). *A multilevel weighted bootstrap procedure to handle sampling weights* [Paper session]. AERA Annual Meeting, Orlando, FL, United States. https://convention2.allacademic.com/one/aera/aera21

### 2020

Lai, M. H. C. (2020, July). *Multilevel bootstrap confidence intervals for standardized effect size* [Paper Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Tse, W. W. & Lai, M. H. C. (2020, July 14-17). *Measurement Invariance for Ordered Categorical Variables: The Necessity of Strict Invariance for Valid Group Comparisons* [Poster Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Zhang, Y. & Lai, M. H. C. (2020, July 14-17). *A Bayesian Region of Measurement Equivalence (ROME) Approach for Establishing Measurement Invariance* [Poster Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Tse, W. W. & Lai, M. H. C. (2020, June 25-26). *Measurement Invariance for Ordered Categorical Variables: The Necessity of Strict Invariance for Valid Group Comparisons* [Conference Session]. A Meeting of Methodologies (AMOM), Vancouver, BC, Canada. [Powerpoint slides](https://mfr.ca-1.osf.io/render?url=https://osf.io/qcfpg/?direct%26mode=render%26action=download%26mode=render)

Lai, M. H. C. (2020, June 2-3). *Constructing bootstrap confidence intervals for standardized effect size in cluster-randomized trials* [Paper Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)

Tse, W. W. & Lai, M. H. C. (2020, June 2-3). *Path to valid means comparisons: The necessity of strict measurement invariance for ordered categorical variables* [Poster Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)

Zhang, Y. & Lai, M. H. C. (2020, June 2-3). *A Bayesian Region of Measurement Equivalence (ROME) Approach for Establishing Measurement Invariance* [Poster Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)
