---
title: "Publications"
description: |
  Recent publications and conference presentations. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r write_refs, include = FALSE, cache = FALSE}
library(RefManageR)
# Grab bibTeX using DOIs, and save to pubs.bib
RefManageR::GetBibEntryWithDOI(
  unique(
    c("10.1080/10705511.2020.1782206", 
      "10.1037/met0000287", 
      "10.1080/00273171.2020.1746902", 
      "10.3102/1076998619843168", 
      "10.1016/j.addbeh.2018.11.029")
  ), 
  temp.file = "pubs.bib", 
  delete.file = FALSE)
# Manual bibTeX for articles without DOIs
ref <- c(
  BibEntry(
    bibtype = "unpublished", 
    key = "lai2021", 
    title = "Adjusting for partial invariance in latent parameter estimation: Comparing forward specification search and approximate invariance methods", 
    author = "Lai, Mark H. C. and Liu, Yuanfang and Tse, Winnie Wing-Yee", 
    year = "2021", 
    journal = "Behavior Research Methods", 
    date = "2021-02-05", 
    pubstate = "Manuscript accepted for publication"
  )
)
WriteBib(bib = ref, 
         file = "pubs.bib", 
         append = TRUE)
```

```{r, load_refs, include = FALSE, cache = FALSE}
library(RefManageR)
# Grab bibTeX using DOIs, and save to pubs.bib
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE, 
           no.print.fields = "url")
myBib <- ReadBib(here::here("pubs.bib"), check = FALSE)
```

## Peer-Reviewed Articles (Selected)

### 2021

```{r, results = "asis"}
myBib[date = "2021"]
```

### 2020

```{r, results = "asis"}
myBib[date = "2020"]
```


### 2019

```{r, results = "asis"}
myBib[date = "2019"]
```

## Conference Presentations

### 2021

Lai, M. H. C. & Hsiao, Y. (2021, April). *Two-stage path analysis with definition variables: A general framework for modeling with categorical indicators* [Structured Poster Session]. AERA Annual Meeting, Orlando, FL, United States. 

### 2020

Lai, M. H. C. (2020, July). *Multilevel bootstrap confidence intervals for standardized effect size* [Paper Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Tse, W. W. & Lai, M. H. C. (2020, July 14-17). *Measurement Invariance for Ordered Categorical Variables: The Necessity of Strict Invariance for Valid Group Comparisons* [Poster Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Zhang, Y. & Lai, M. H. C. (2020, July 14-17). *A Bayesian Region of Measurement Equivalence (ROME) Approach for Establishing Measurement Invariance* [Poster Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Tse, W. W. & Lai, M. H. C. (2020, June 25-26). *Measurement Invariance for Ordered Categorical Variables: The Necessity of Strict Invariance for Valid Group Comparisons* [Conference Session]. A Meeting of Methodologies (AMOM), Vancouver, BC, Canada. [Powerpoint slides](https://mfr.ca-1.osf.io/render?url=https://osf.io/qcfpg/?direct%26mode=render%26action=download%26mode=render)

Lai, M. H. C. (2020, June 2-3). *Constructing bootstrap confidence intervals for standardized effect size in cluster-randomized trials* [Paper Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)

Tse, W. W. & Lai, M. H. C. (2020, June 2-3). *Path to valid means comparisons: The necessity of strict measurement invariance for ordered categorical variables* [Poster Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)

Zhang, Y. & Lai, M. H. C. (2020, June 2-3). *A Bayesian Region of Measurement Equivalence (ROME) Approach for Establishing Measurement Invariance* [Poster Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)
