---
title: "Publications"
description: |
  Recent publications and conference presentations. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r write_refs, include = FALSE, cache = FALSE}
library(RefManageR)
# Grab bibTeX using DOIs, and save to pubs.bib
RefManageR::GetBibEntryWithDOI(
  unique(
    c(readLines("pub_doi.txt"))
  ), 
  temp.file = "pubs.bib", 
  delete.file = FALSE)
# Manual bibTeX for articles without DOIs
if (length(readLines("pub_nodoi.bib")) > 1 ||
    readLines("pub_nodoi.bib") != "") {
  ref <- RefManageR::ReadBib("pub_nodoi.bib")
  WriteBib(bib = ref, 
           file = "pubs.bib", 
           append = TRUE)
}
```

```{r, load_refs, include = FALSE, cache = FALSE}
library(RefManageR)
# Read "pubs.bib"
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE, 
           no.print.fields = "url")
myBib <- ReadBib(here::here("pubs.bib"), check = FALSE)
```

## Peer-Reviewed Articles (Selected)

```{r, results = "asis"}
for (yeari in as.character(2022:2019)) {
  cat(paste("\n###", yeari), "\n")
  print(myBib[date = yeari])
}
```

## Conference Presentations

### 2023

Zhang, G., & Lai, M. H. C. (2023, July 25-28). A Two-Stage Path Analysis Approach to Model Interaction Effects for Congeneric Measures [Oral presentation]. European Congress of Methodology (EAM), Ghent, Belgium. https://eam2023.ugent.be/images/eam2023_abstracts_book.pdf

Zhang, Y., & Lai, M. H. C. (2023, July 25-28). Alignment with Bayesian Region of Measurement Equivalence (ABROME) approach for multiple groups comparisons [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), College Park, Maryland, United States. https://modeling.uconn.edu/2023-preliminary-program/

Tse, W. W. Y., & Lai, M. H. C. (2023, July 25-28). Addressing publication bias and uncertainty for power analysis: A hybrid classical-Bayesian approach [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), College Park, Maryland, United States. https://modeling.uconn.edu/2023-preliminary-program/

Tse, W. W. Y., & Lai, M. H. C. (2023, July 25-28). Addressing publication bias and uncertainty for power analysis: A hybrid classical-Bayesian approach [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), College Park, Maryland, United States. https://modeling.uconn.edu/2023-preliminary-program/

Liu, Y., & Lai, M. H. C. (2023, July 25-28). Examining SEM Trees for Investigating Measurement Invariance Concerning Multiple Violators [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), College Park, Maryland, United States. https://modeling.uconn.edu/2023-preliminary-program/

Blozis, S., & Lai, M. H. C. (2023, July 25-28). A two-stage approach to a latent variable mixed-effects location scale model [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), College Park, Maryland, United States. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps2023-abstracts.pdf

Blozis, S., & Lai, M. H. C. (2023, July 25-28). A two-stage approach to a latent variable mixed-effects location scale model [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), College Park, Maryland, United States. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps2023-abstracts.pdf

Liu, Y., & Lai, M. H. C. (2023, June 26-28). Examining SEM Trees for Investigating Measurement Invariance Concerning Multiple Violators [Paper session]. Modern Modeling Methods Conference, Storrs, CT, United States. https://modeling.uconn.edu/2023-preliminary-program/

Ji, Y., Luo, W., Lai, M. H. C., Kwok. O., & Yoon, M. (2023, April 13-16). An Eigenvalues-Based Approaches to Identifying the Optimal Number of Factors in Multilevel Exploratory Factor Analysis [Roundtable presentation]. AERA Annual Meeting, Chicago, IL, United States. https://convention2.allacademic.com/one/aera/aera23

Lai, M. H. C., Zhang, Y., & Ji, F. (2023, April 13-16). An Empirical Bayes Cluster-Mean Approach to Correct for Sampling Error in Between-Cluster Effects [Poster presentation]. AERA Annual Meeting, Chicago, IL, United States. https://convention2.allacademic.com/one/aera/aera23

Tse, W. W. Y., & Lai, M. H. C. (2023, April 13-16). Addressing Uncertainty in Power Analysis for Designing Cluster Randomized Trials. [Paper session]. AERA Annual Meeting, Chicago, IL, United States. https://convention2.allacademic.com/one/aera/aera23

### 2022

Lai, M. H. C., Tse, W. W. Y., Zhang, G., Li, Y., & Hsiao, Y. Y. (2022, September 21-24). *Correcting for unreliability and partial invariance: A two-stage path analysis approach* [Symposium presentation]. Society for Research on Educational Effectiveness 2022 Conference, Arlington, VA, United States. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps2022-version-7-abstract.pdf

Zhang, G., & Lai, M. H. C. (2022, July 11-15). *A systematic review of measurement invariance research of the CES-D across gender: calculation and report of effect size* [Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), Bologna, Italy. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps2022-version-7-abstract.pdf

Tse, W. W., & Lai, M. H. C. (2022, July 11-15). *Incorporating information from historical data for power analysis: A hybrid classical-Bayesian approach for multilevel studies *[Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), Bologna, Italy. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps2022-version-7-abstract.pdf

Zhang, Y., & Lai, M. H. C. (2022, July 11-15). *Bayesian region of measurement equivalence (ROME) approach with alignment *[Oral presentation]. Annual Meeting of the Psychometric Society (IMPS), Bologna, Italy. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps2022-version-7-abstract.pdf

Tse, W. W., & Lai, M. H. C. (2022, June 27-29). *Incorporating Uncertainty in power analysis to more realistically determine sample size* [Workshop]. Meeting of the Society for the Improvement of Psychological Science (SIPS), Victoria, B. C., Canada. https://docs.google.com/document/d/1Rka3f0ZY9BBZVIlZoVINrpyaJYeitpq1yKXBLukeMng

Zhang, Y., & Lai, M. H. C. (2022, April 21-26). *Evaluating standard error estimators and resampling method for multilevel models on small samples with heteroscedasticity and unbalanced cluster sizes *[Poster presentation]. AERA Annual Meeting, San Diego, CA, United States. https://convention2.allacademic.com/one/aera/aera22

Palardy, G. J., Zhang, Y., & Lai, M. H. C. (2022, April 21-26). *Measurement invariance testing via the alignment method: Intersectional grouping and multiple cohorts* [oral presentation]. AERA Annual Meeting, San Diego, CA, United States. https://convention2.allacademic.com/one/aera/aera22

### 2021

Zhang, G., Yue, H., & Lai, M. H. C. (2021, July 19-23). *A systematic review of measurement invariance research of the CES-D across gender* [Poster presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Zhang, Y. & Lai, M. H. C. (2021, July 19-23). *Classification Accuracy of multidimensional tests: Quantifying the impact of noninvariance* [Oral presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Tse, W. W. & Lai, M. H. C. (2021, July 19-23). *Incorporating uncertainty in design parameters: A hybrid classical-Bayesian power analysis approach for two-level cluster randomized trials* [Oral presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Lai, M. H. C., Ji, F., & Chen, S. (2021, July 19-23). *Obtaining reliability for daily diary data using multilevel factor analysis* [Oral presentation]. Virtual conference of the Psychometric Society (IMPS) due to covid-19. https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2021_virtual_schedule.pdf

Hsiao, Y., Kwok, O., & Lai, M. H. C. (2021, April 8-12). *Improving estimation accuracy of the tnteraction effects in models with composite scores* [Structured poster session]. AERA Annual Meeting, Orlando, FL, United States. https://convention2.allacademic.com/one/aera/aera21

Lai, M. H. C. & Hsiao, Y. (2021, April). *Two-stage path analysis with definition variables: A general framework for modeling with categorical indicators* [Structured Poster Session]. AERA Annual Meeting, Orlando, FL, United States. 

Wen, L., & Lai, M. H. C. (2021, April 8-12). *A multilevel weighted bootstrap procedure to handle sampling weights* [Paper session]. AERA Annual Meeting, Orlando, FL, United States. https://convention2.allacademic.com/one/aera/aera21

### 2020

Lai, M. H. C. (2020, July). *Multilevel bootstrap confidence intervals for standardized effect size* [Paper Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Tse, W. W. & Lai, M. H. C. (2020, July 14-17). *Measurement Invariance for Ordered Categorical Variables: The Necessity of Strict Invariance for Valid Group Comparisons* [Poster Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Zhang, Y. & Lai, M. H. C. (2020, July 14-17). *A Bayesian Region of Measurement Equivalence (ROME) Approach for Establishing Measurement Invariance* [Poster Session]. Annual Meeting of the Psychometric Society (IMPS), College Park, MD, United States. [Schedule](https://www.psychometricsociety.org/sites/main/files/file-attachments/imps_2020_virtual_schedule_g1.pdf?1594692842)

Tse, W. W. & Lai, M. H. C. (2020, June 25-26). *Measurement Invariance for Ordered Categorical Variables: The Necessity of Strict Invariance for Valid Group Comparisons* [Conference Session]. A Meeting of Methodologies (AMOM), Vancouver, BC, Canada. [Powerpoint slides](https://mfr.ca-1.osf.io/render?url=https://osf.io/qcfpg/?direct%26mode=render%26action=download%26mode=render)

Lai, M. H. C. (2020, June 2-3). *Constructing bootstrap confidence intervals for standardized effect size in cluster-randomized trials* [Paper Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)

Tse, W. W. & Lai, M. H. C. (2020, June 2-3). *Path to valid means comparisons: The necessity of strict measurement invariance for ordered categorical variables* [Poster Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)

Zhang, Y. & Lai, M. H. C. (2020, June 2-3). *A Bayesian Region of Measurement Equivalence (ROME) Approach for Establishing Measurement Invariance* [Poster Session]. Modern Modeling Methods Conference (MMM), Storrs, CT, United States. (Conference canceled)
