---
title: Correlation Attenuation for Categorical Variables
author: Gengrui (Jimmy) Zhang
date: '2023-04-19'
output:
  distill::distill_article:
    self_contained: false
    toc: true
categories:
  - statistics
  - correlation
  - categorical
draft: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r loading packages, include = FALSE}
# HL: I took out packages not used
library(dplyr)
library(MASS)
library(mnormt)
# HL: The seed should be set for random number generation
set.seed(2229)
```

# An Intro to Correlation Attenuation

Correlation is the degree to which two variables associate with one another. The correlation formula between two random variables (i.e., X and Y) is:


$$\rho(x,y) = \frac{COV(X, Y)}{\sigma_{X}\sigma_{Y}},$$

where $\sigma$ is the standard deviation. 

When one of the variable is categorized into dichotomous or categorical variables, the correlation $\rho(X,Y)$ will usually be attenuated due to loss of information. 

## An Example of Attenuated Correlation for Dichotomous Variable

Say $X$ and $Y^*$ have a correlation of .5 (i.e., $\rho[X, Y^*] = .5$), $Y^*$ is dichotomized into $Y$ so that 30% of $Y$ is 0 and 70% of $Y$ is 1. What is the correlation between $X$ and $Y$ now?

Let's simulate a dataset to see this attenuation:

```{r}
# Set correlation between X and Y* to 0.5
rho <- 0.5

# Assume X and Y*~N(0,1) for now
sd_x <- 1
sd_y <- 1
cov_xy <- rho*sd_x*sd_y

# Simulate correlated X and Y*
df <- as.data.frame(
  mvrnorm(n = 1e4,
          mu = c(0, 0),
          Sigma = matrix(c(sd_x^2, cov_xy,
                           cov_xy, sd_y^2),
                         ncol = 2))
  )
names(df) <- c("X", "Y*")

# Manually dichotomize Y* to 0 and 1
df <- df %>%
  mutate(Y = ifelse(`Y*` > qnorm(0.7, mean(`Y*`), sd(`Y*`)), 1, 0))

# Show proportion of Y
knitr::kable(table(df$Y)/nrow(df),
             col.names = c("Label", "Proportion"),
             align = "c")

# Show correlations between X and Y*, and X and Y
knitr::kable(
  cbind(cor(df$X, df$`Y*`), cor(df$X, df$Y)),
  col.names = c("$\\rho_{(X, Y*)}$", "$\\rho_{(X, Y)}$") 
)
```

<!-- HL: Missing discussion on only needing the correlation, not the covariance -->

From this example, we can see the correlation is attenuated when one of the continuous variable is dichotomized. According to the correlation formula and expectation of covariance formula, we can derive the attenuation factor due to categorization. Note that the value of dichotomozing $Y^*$ for desired proportion is called `threshold`. 


$$\text{Attenuation Factor} = \frac{COV(X, Y)}{COV(X, Y^*)}*\sqrt{\frac{\sigma^2_{Y^*}}{\sigma^2_{Y}}},$$



$$\text{Attenuation Factor} = \frac{E(XY) - E(X)E(Y)}{E(XY^*) - E(X)E(Y^*)}*\sqrt{\frac{\sigma^2_{Y^*}}{\sigma^2_{Y}}}$$


Thresholds are only dependent on the variable being categorized (i.e., $Y^*$ in this case). To calculate the expectation of $XY$ and $XY^*$ in the formula above, we need to use bivariate normal distribution:

<!-- HL: Unclear how the formula and discussion below relates to E(XY) -->

$$f_{x,y}(x,y) = \frac{1}{2\pi\sigma_{x}\sigma_{y}\sqrt{1 - \rho^2}}*e^{-\frac{1}{2(1 - \rho^2)}*\left[(\frac{x - \mu_{x}}{\sigma_{x}})^2 + (\frac{y - \mu_{y}}{\sigma_{y}})^2 - 2\rho\frac{(x - \mu_{x})(y - \mu_{y})}{\sigma_{x}\sigma_{y}}\right]}$$

<!-- HL: Do you use the density function or cumulative probability? Please clarify as the above formula is density, which does not require integration. -->

We can code this formula into a `R` function and use it later. 

```{r bivariate normal function}
# helper function for integrating bivariate normal density
pbnorm <- function(lo1, up1, lo2, up2, mu1, mu2, sigma1, sigma2, rho) {
  cubature::cuhre(
    function(arg) {
      y1 <- arg[1]
      y2 <- arg[2]
      # bivariate normal density function
      ((1/(2*pi*sigma1*sigma2*sqrt(1 - rho^2))) *
          exp(- ((y1-mu1)^2/sigma1^2 - 2*rho*(y1-mu1)*(y2-mu2)/(sigma1*sigma2) +
                   (y2-mu2)^2/sigma2^2) / (2*(1 - rho^2))))*y1
    },
    lowerLimit = c(lo1, lo2), upperLimit = c(up1, up2)
  )$integral
}
```

Now we can use the derived formula instead of simulated dataset to calculated attenuated $R^2$:

```{r}
# Analytic calculation
thres <- qnorm(0.3)
var_ystar <- 1
var_y <- 0.7*(1-0.7)  # HL: This has not been discussed
# HL: If you're integrating one of the variables in the bivariate normal out, you just need the marginal distribution (Y* in this case)
# HL: You can verify that the one below is just the mean of a truncated normal and does not depend on rho. This can be discussed.
# attenuation_bi <- pbnorm(-Inf, Inf, thres, Inf, 0, 0, sd_x, sd_y, rho)/rho * sqrt(var_ystar/var_y)
attenuation_bi <- dnorm(thres) * sqrt(var_ystar / var_y)
cor_xy_bi <- attenuation_bi * rho

# Simulated results
lat_cor <- cor(df$X, df$`Y*`)
obs_cor <- cor(df$X, df$Y)

att_fac <- (cov(df$X, df$Y)/cov(df$X, df$`Y*`))*sqrt(var(df$`Y*`)/var(df$Y))
cal_cor <- att_fac*lat_cor
```

Then we can compare the results from analytic calculation and simulated results:

```{r}
summary_1 <- round(c(rho, attenuation_bi, cor_xy_bi, att_fac, cal_cor), 3)
names(summary_1) <- c("Correlation_XY*", "Attenuation_Formula", 
                      "Correlation_Formula", "Attenuation_Data", 
                      "Correlation_Data")
knitr::kable(summary_1,
             align = "c",
             col.names = " ")
```

## An Example of Attenuated Correlation for Categorical Variable with Three Thresholds

Given $Y^*$ is discretized into $Y$ with 4 categories (ie., 50% is 0, 30% is 1, 10% is 2, 10% is 3), what is the correlation between $X$ and $Y$?

```{r}
# Assuming X and Y* ~ N(0,1)
# for standard bivariate normal distribution, E(XY*) = rho
rho <- 0.5
var_ystar <- 1

thres_1 <- qnorm(0.5)
thres_2 <- qnorm(0.5 + 0.3)
thres_3 <- qnorm(0.5 + 0.3 + 0.1)

p_less_than_thres1 <- pnorm(thres_1)
p_thres1_thres2 <- pnorm(thres_2) - pnorm(thres_1)
p_thres2_thres3 <- pnorm(thres_3) - pnorm(thres_2)
p_larger_than_thres3 <- pnorm(thres_3, lower.tail = F)

e_y2 <- 0*p_less_than_thres1 + 1^2*p_thres1_thres2 + 2^2*p_thres2_thres3 + 3^2*p_larger_than_thres3
e_y <- 1*p_thres1_thres2 + 2*p_thres2_thres3 + 3*p_larger_than_thres3
var_y <- e_y2 - e_y^2
# HL: Verify you can simplify to the following (based on the binary results):
attenuation_cat <- 1 * (dnorm(thres_1) - dnorm(thres_2)) + 2 * (dnorm(thres_2) - dnorm(thres_3)) + 3 * dnorm(thres_3) * sqrt(var_ystar / var_y)
# attenuation_cat <- (0*pbnorm(-Inf, Inf, -Inf, thres_1, 0, 0, 1, 1, 0.5) + 
#   1*pbnorm(-Inf, Inf, thres_1, thres_2, 0, 0, 1, 1, 0.5) + 
#   2*pbnorm(-Inf, Inf, thres_2, thres_3, 0, 0, 1, 1, 0.5) + 
#   3*pbnorm(-Inf, Inf, thres_3, Inf, 0, 0, 1, 1, 0.5))/rho * sqrt(var_ystar/var_y)
cor_xy_cat <- attenuation_cat*rho
```

# Verification with simulated data

```{r}
rho <- 0.5
sd_x <- 1
sd_y <- 1
cov_xy <- rho*sd_x*sd_y

df3 <- as.data.frame(mvrnorm(n = 1e4,
                             mu = c(0, 0),
                             Sigma = matrix(c(sd_x^2, cov_xy,
                                              cov_xy, sd_y^2),
                                            ncol = 2)))


names(df3) <- c("y1", "y2")

df3 <- df3 %>%
  mutate(y2_mul = ifelse(y2 < qnorm(0.5, mean(df3$y2), sd(df3$y2)), 0,
                         ifelse(qnorm(0.5, mean(df3$y2), sd(df3$y2)) < y2 & y2 < qnorm(0.5 + 0.3, mean(df3$y2), sd(df3$y2)), 1,
                                ifelse(qnorm(0.5 + 0.3, mean(df3$y2), sd(df3$y2)) < y2 & y2 < qnorm(0.5 + 0.3 + 0.1, mean(df3$y2), sd(df3$y2)), 2,
                                       ifelse(y2 > qnorm(0.5 + 0.3 + 0.1, mean(df3$y2), sd(df3$y2)), 3, NA)))))

lat_cor <- cor(df3$y1, df3$y2)
obs_cor <- cor(df3$y1, df3$y2_mul)

att_fac <- (cov(df3$y1, df3$y2_mul)/cov(df3$y1, df3$y2))*sqrt(var(df3$y2)/var(df3$y2_mul))
cal_cor <- att_fac*lat_cor
```

```{r}
summary_2 <- round(c(rho, attenuation_cat, cor_xy_cat, att_fac, cal_cor), 3)
names(summary_2) <- c("Correlation_XY*", "Attenuation_Formula", 
                      "Correlation_Formula", "Attenuation_Data", 
                      "Correlation_Data")
knitr::kable(summary_2,
             align = "c",
             col.names = " ")
```

# Reasoning of Generalization to X and Y* with Any Means and Variances

It is highly probable that $X$ and $Y*$ do not follow a standard normal distribution in practical research. We would like to prove that the attenuation of correlation is generalizeable to $X$ and $Y^*$ with any means and variances when $Y^*$ is categorized with any numbers of categories.

<!-- HL: Unclear why there two 'c's below -->

let's say $Y^{*}$ is categorized to Y (c = 4; c = \[0, 1, 2, 3\]),

$$E(X,Y) = 
  \begin{cases}
    0, & \text{if } Y^{*} \le \tau_{1} \\  
    E(X | Y = 1), & \text{if } \tau_{1} \le Y^{*} \le \tau_{2} \\
    E(X | Y = 2), & \text{if } \tau_{2} \le Y^{*} \le \tau_{3} \\
    E(X | Y = 3), & \text{if } Y^{*} > \tau_{3}
  \end{cases}$$


Take one category as one example, for $E(X | Y = 1)$ with $\tau_{1} \le Y^{*} \le \tau_{2}$ and $c = 1$: $E(X | Y = 1) = \int_{-\infty}^{\infty} \int_{\tau_{1}}^{\tau_{2}} \text{x} y^{*} f_{(x, y^{\ast})} d_{x} d_{y^{*}}$

Now the distributions of X and Y are dependent on their mean and variance. We can use the z-scores to subsititute limits of integrals.

Let $z_{x} = \frac{x - \mu_{x}}{\sigma_{x}}$ and $z_{y^{*}} = \frac{y^{*} - \mu_{y^{*}}}{\sigma_{y^{*}}}$, then,


$$f_{x,y^{*}}(x,y^{*}) = \frac{1}{2\pi\sigma_{x}\sigma_{y^{*}}\sqrt{1 - \rho^2}}*e^{-\frac{1}{2(1 - \rho^2)}*[z_{x}^2 + z_{y^{*}}^2 - 2\rho z_{x} z_{y^{*}}]},$$

and transform the formula into: 


$$E(X | Y = 1) = \int_{-\infty}^{\infty}\int_{\tau_{1}}^{\tau_{2}}xy^{*}\frac{d_{x}d_{y^{*}}}{2\pi\sigma_{x}\sigma_{y^{*}}\sqrt{1 - \rho^2}}*e^{-\frac{1}{2(1 - \rho^2)}*[z_{x}^2 + z_{y^{*}}^2 - 2\rho z_{x} z_{y^{*}}]}$$

Because of the property of derivation,


$$\frac{d(x)}{\sigma_{x}} = d(\frac{x - \mu_{x}}{\sigma_{x}}) = d(z_{x}),$$

and it is the same for $d(z_{y^{*}})$.

Thus, the equation of $E(X | Y = 1)$ becomes:


$$E(X | Y = 1) = \int_{-\infty}^{\infty}\int_{\frac{\tau_{1} - \mu_{y^{*}}}{\sigma_{y^{*}}}}^{\frac{\tau_{2} - \mu_{y^{*}}}{\sigma_{y^{*}}}}xy^{*}\frac{1}{2\pi\sqrt{1 - \rho^2}}*e^{-\frac{1}{2(1 - \rho^2)}*[z_{x}^2 + z_{y^{*}}^2 - 2\rho z_{x} z_{y^{*}}]}d(z_{x})d(z_{y^{*}})$$

The "new" values of limits, e.g., $\frac{\tau_{1} - \mu_{y^{*}}}{\sigma_{y^{*}}}$, are linear tranformed using the mean and variance of $Y^{*}$. It means that no matter how threshold values change due to mean and variance of $Y^{*}$, we can always z-tranform them back so that X and $Y^{*}$ always follow a standard bivariate normal distribution. In other words, as long as we know the threshold values and proportion of categories of the categorized variable, and X and $Y^{*}$ follow normal distributions, we should be able to compute the attenuated $R^2$ no matter the mean and variance of $Y^{*}$.

# Verification with simulated data (random mean and variance)

Now it's time to verify if our reasoning works with any means and variances for dichotomous $Y$ and categorical $Y$.

```{r}
rho <- 0.5
sd_x <- rnorm(1, 1, 0.5)
sd_y <- rnorm(1, 1.5, 0.3)
cov_xy <- rho*sd_x*sd_y

df2 <- as.data.frame(mvrnorm(n = 1e7,
                              mu = c(rnorm(1, 10, 2.1), rnorm(1, 8, 1.1)),
                              Sigma = matrix(c(sd_x^2, cov_xy,
                                               cov_xy, sd_y^2),
                                             ncol = 2)))

names(df2) <- c("y1", "y2")
df2 <- df2 %>%
  mutate(y2_cat = ifelse(y2 > qnorm(0.7, mean(df2$y2), sd(df2$y2)), 1, 0))

lat_cor <- cor(df2$y1, df2$y2)
obs_cor <- cor(df2$y1, df2$y2_cat)

att_fac <- (cov(df2$y1, df2$y2_cat)/cov(df2$y1, df2$y2))*sqrt(var(df2$y2)/var(df2$y2_cat))
cal_cor <- att_fac*lat_cor
```

```{r}
summary_3 <- round(c(rho, attenuation_bi, cor_xy_bi, att_fac, cal_cor), 3)
names(summary_3) <- c("Correlation_XY*", "Attenuation_Formula", 
                      "Correlation_Formula", "Attenuation_Data", 
                      "Correlation_Data")
knitr::kable(summary_3,
             align = "c",
             col.names = " ")
```

```{r}
rho <- 0.5
sd_x <- rnorm(1, 1, 0.5)
sd_y <- rnorm(1, 1.5, 0.3)
cov_xy <- rho*sd_x*sd_y

df3 <- as.data.frame(mvrnorm(n = 1e7,
                             mu = c(rnorm(1, 10, 2.1), rnorm(1, 8, 1.1)),
                             Sigma = matrix(c(sd_x^2, cov_xy,
                                              cov_xy, sd_y^2),
                                            ncol = 2)))


names(df3) <- c("y1", "y2")

df3 <- df3 %>%
  mutate(y2_mul = ifelse(y2 < qnorm(0.5, mean(df3$y2), sd(df3$y2)), 0,
                         ifelse(qnorm(0.5, mean(df3$y2), sd(df3$y2)) < y2 & y2 < qnorm(0.5 + 0.3, mean(df3$y2), sd(df3$y2)), 1,
                                ifelse(qnorm(0.5 + 0.3, mean(df3$y2), sd(df3$y2)) < y2 & y2 < qnorm(0.5 + 0.3 + 0.1, mean(df3$y2), sd(df3$y2)), 2,
                                       ifelse(y2 > qnorm(0.5 + 0.3 + 0.1, mean(df3$y2), sd(df3$y2)), 3, NA)))))

lat_cor <- cor(df3$y1, df3$y2)
obs_cor <- cor(df3$y1, df3$y2_mul)

att_fac <- (cov(df3$y1, df3$y2_mul)/cov(df3$y1, df3$y2))*sqrt(var(df3$y2)/var(df3$y2_mul))
cal_cor <- att_fac*lat_cor
```

```{r}
summary_4 <- round(c(rho, attenuation_cat, cor_xy_cat, att_fac, cal_cor), 3)
names(summary_4) <- c("Correlation_XY*", "Attenuation_Formula", 
                      "Correlation_Formula", "Attenuation_Data", 
                      "Correlation_Data")
knitr::kable(summary_4,
             align = "c",
             col.names = " ")
```

It seems that the comparison of attenuated $R^2$ values calculated by the formula and from the simulated results are highly similar. 

